### 2015-03-09 doc/thesis/system.tex
More improvements to the text about typing rules

Our core limits control flow to if and while statements;
it has explicit type annotations, explicit scope for variables,
explicit method declarations, and explicit method calls.
Here is a list of features that are not present in our reduced core:
\begin{itemize}
\item labels and goto statements are not present because they
are difficult to handle along with our simplified form of
\emph{flow typing}, and they are out of scope for now;
\item explicit blocks are not present because we are already using
explicit scope for variables;
\item other loop structures such as repeat-until, numeric for,
and generic for are not present because we can use while to
express them;
\item fields in the table constructor are reduced to a single
form because we can use this form to express the missing forms; 
\item arithmetic operators are reduced to addition only
because the other operators have similar typing rules;
\item relational operators are reduced to equality and less than
because the other operators have similar typing rules;
\item bitwise operators are reduced to bitwise and only
because the other operators have similar typing rules.
\end{itemize}

@@

Our reduced core includes two statements for declaring local variables,
one with and another without type annotations.
While we use the former to show how our type system handles the declaration
of annoatated variables, we use the latter to show how our type system
handles the declaration of unannotated variables through local type inference
and also the introduction of projection types.

Our reduced core also includes a truncation operator $\lfloor \rfloor$ for
function applications, method applications, and the vararg expression.
We use $\lfloor a \rfloor_{0}$ to denote function and method applications
that produce no value because they appear as statements.
We use $\lfloor me \rfloor_{1}$ to denote function applications,
method applications, and vararg expressions that produce only one value,
even if they return mutiple values.

@@

As an example, lets assume that $x$ and $y$ are variables in the
environment with types $\Integer$ and $\String$, the assignment
\[
x, y = 1, ``foo"
\]
type checks through the following rules:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env_{i}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       n = |\;\overline{e}\;| \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1})}
      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env, \penv \vdash l_{i}:T_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\end{array}
\]

@@

The previous example type checks through these rules because
\[
1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}
\]

As another example, lets assume that $x$, $y$, and $z$ are variables in
the environment with types $\Integer$, $\String$, and $\String \cup \Nil$,
the assignment
\[
x, y, z = 1, ``foo"
\]
type checks because
\[
\Integer \times \String \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
\]

Conversely, the assignment
\[
x = 1, ``foo"
\]
type checks because
\[
\Integer \times \String \times \Nil{*} \lesssim \Integer \times \Value{*}
\]

@@

Assuming that $f$ is a local function in the environment, and that $f$ has type
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*} \rightarrow \Integer{*}$,
the function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]
and the function call
\[
f(``foo",1,2,3)
\]
also type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]

@@

When our type system type checks an expression list,
it always includes $\Nil{*}$ in the end of the type of this expression list
if its type does not end in a variadic type.
This behavior preserves the semantics of Lua on replacing missing values,
and it is necessary when we omit optional parameters in a function call,
like the previous example showed.

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
We made this restriction because our type system does not keep
track of aliases to table fields.
This means that allowing \emph{unique} and \emph{open} table
types to appear in the type of the value would allow the
creation of unsafe aliases.

After discussing the typing rules of the table constructor,
we start the discussion of the rules that define the most
unusual feature of our type system: the refinement of table types.
The first kind of refinement allows programmers to add new
fields to \emph{unique} or \emph{open} table types through
field assignment.
For instance, in Section \ref{sec:tables} we presented the
following example:
\begin{verbatim}
    local person = {}
    person.firstname = "Lou"
    person.lastname = "Reed"
\end{verbatim}

We translate this example to our reduced core as follows:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; person:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$person[``firstname"] \; {<}\String{>} = ``Lou";$}\\
& \multicolumn{1}{l}{$person[``lastname"] \; {<}\String{>} = ``Reed"$}
\end{tabular}
\end{center}

@@

We use the refinement of table types to handle the declaration of
new global variables.
In Lua, the assignment \texttt{v = v + 1} translates to
\texttt{\string_ENV["v"] = \string_ENV["v"] + 1} when \texttt{v}
is not a local variable, where \texttt{\string_ENV} is a table
that stores the global environment.
For this reason, Typed Lua treats accesses to global variables as field accesses
to an \emph{open} table in the top-level scope.
In the following examples we assume that $\string_ENV$ is in the
environment and has type $\{\}_{open}$. For instance,
\[
\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``y"] \; {<}\Integer{>} = 1
\]
uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
Therefore, after these field assignments $\string_ENV$ has type
$\{``x":\String, ``y":\Integer\}_{open}$.

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this behavior:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env, \penv \vdash id:close(T_{2}), \env[id \mapsto open(T_{1})]}
\end{array}
\;
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env, \penv \vdash id:fix(T_{2}), \env[id \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

Both rules use the auxiliary function \emph{read} because they may be
accessing an identifier that is bound to a filter or projection type.
As we mentioned in Section \ref{sec:unions}, our type system includes
a small set of \texttt{type} predicates that allow programmers to
discriminate union types, and our type system uses filter and projection
types in the definition of these predicates to handle the discrimination
of unions of first-level and second-level types.
We can define \emph{read} as follows:
\begin{align*}
read(\penv, \phi(T_{1},T_{2})) & = T_{2}\\
read(\penv, \pi_{i}^{x}) & = proj(\penv(x), i)\\
read(\penv, T) & = T
\end{align*}

@@

The following example does not type check, as aliasing $a$ produces
the type $\{``foo":\Integer\}_{closed}$ that is not a subtype of
$\{``foo":\Integer \cup \Nil\}_{closed}$, the type of $b$:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{``foo":\Integer\}_{unique} = \{ [``foo"] = 5 \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{closed} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$b[``foo"] = \mathbf{nil}$}
\end{tabular}
\end{center}

As another example,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
does not type check, as aliasing $a$ produces the type $\{\}_{closed}$
that is not a subtype of $\{\}_{open}$, the type of $b$.
Our type system has this behavior to warn programmers about
potential unsafe behaviors after this kind of alias.
In this example, it is unsafe to add the field $``x"$ to $b$,
as it changes the value that is stored in the field $``x"$ of $a$.

@@

As an example, rule \textsc{T-FUNCTION3} prevents the following unsafe example to type check:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Void \rightarrow \Integer \times \Void =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Void$}\\
& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])$}
\end{tabular}
\end{center}

@@

First, our type system uses rule \textsc{T-EXPLIST6} for type checking
$idiv(1, 2)$:
\[
\begin{array}{c}
\mylabel{T-EXPLIST6}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash me:S, \env_{2}\\
       S = T_{1} \times ... \times T_{n} \times \Void \sqcup T_{1}' \times ... \times T_{n}' \times \Void
       \end{array}}
      {\env_{1}, \penv \vdash me:\pi_{1}^{x} \times ... \times \pi_{n}^{x} \times \Nil{*}, \env_{2}, (x,S)}
\end{array}
\]

Rule \textsc{T-EXPLIST6} uses the auxiliary relation
$\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
This relation means that given a type environment $\env_{1}$ and
a projection environment $\penv$, we can check that an expression
list $el$ has type $S_{1}$ and produces a new type environment $\env_{2}$
and produces a pair $(x,S_{2})$.
This pair means that the last expression of an expression list $el$
produces a union of second-level types $S_{2}$ that sould be bound
to a variable $x$ in the projection environment,
as the resulting type of this expression is a tuple of projection
types $\pi_{i}^{x}$.
In our example, we can check that $idiv(1, 2)$ has type
$\pi_{1}^{x} \times \pi_{2}^{x} \times \Nil{*}$ and produces the pair
$(x,\Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void)$.

Our type system uses the pair $(x,S_{2})$ to allow rule
\textsc{T-LOCAL2} to bound a union of second-level types $S_{2}$
to a variable $x$ in $\penv$:
\[
\begin{array}{c}
\mylabel{T-LOCAL2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash el:S_{1}, \env_{2}, (x,S_{2})\\
       \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)], \penv[x \mapsto S_{2}] \vdash s, \env_{3} \;\;\;
       n = |\;\overline{id}\;|  
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{3} - \{\overline{id}\}}
\end{array}
\]

This rule uses the auxiliary function \emph{infer} to get the
first-level types of each variable that should be introduced in
the type environment for type checking $s$.
In our example, $q$ gets type $\pi_{1}^{x}$ and $r$ gets type
$\pi_{2}^{x}$.
After type checking the statement $s$, rule \textsc{T-LOCAL} produces a
new environment without the variables that it introduced before type checking $s$.

Introducing a variable $x$ in the projection environment allows our
type system to discriminate projection types $\pi_{i}^{x}$.
The rule \textsc{T-IF3} shows the case where our type system
discriminates a projection type based on the tag \texttt{nil}:
\[
\begin{array}{c}
\mylabel{T-IF3}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
       \env_{4} = openset(closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\]

Rule \textsc{T-IF3} uses the auxiliary functions \emph{fpt} and \emph{gpt}
to filter a projection $x$, affecting all variables that bind to the same projection.
For instance, our previous example type checks through rule \textsc{T-IF3},
because it makes the rule \textsc{T-IF3} use the function call
\[
fpt(\Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void, \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Integer \times \Integer \times \Void$ inside the $\mathbf{if}$ branch,
and the function call
\[
gpt(\Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void, \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Nil \times \String \times \Void$ inside the $\mathbf{else}$ branch.
Thus, reading $q$ and $r$ projects $\pi_{1}^{x}$ to $\Integer$ and
$\pi_{2}^{x}$ to $\Integer$ inside the $\mathbf{if}$ branch,
but it projects $\pi_{1}^{x}$ to $\Nil$ and $\pi_{2}^{x}$ to $\String$
inside the $\mathbf{else}$ branch.
Outside the condition, $q$ and $r$ use the original projection, that is,
they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.

>>>>

Our core limits control flow to if and while statements;
it has explicit type annotations, explicit scope for variables,
explicit method declarations, and explicit method calls.
Here is a list of features that are not present in our reduced core:
\begin{itemize}
\item labels and goto statements are not present because they
are difficult to handle along with our simplified form of
\emph{flow typing}, and they are out of scope for now;
\item explicit blocks are not present because we are already using
explicit scope for variables;
\item other loop structures such as repeat-until, numeric for,
and generic for are not present because we can use while to
express them;
\item fields in the table constructor are reduced to a single
form because we can use this form to express the missing forms; 
\item arithmetic operators are reduced to addition only
because the other operators have similar typing rules;
\item relational operators are reduced to equality and less than
because inequality has similar typing rules to equality and
other relational operators have similar typing rules to less than;
\item bitwise operators are reduced to bitwise and only
because the other operators have similar typing rules.
\end{itemize}

@@

Our reduced core includes two statements for declaring local variables,
one with and another without type annotations.
While we use the former to formalize how our type system handles the declaration
of annotated variables, we use the latter to formalize how our type system
handles the declaration of unannotated variables through local type inference
and also the introduction of projection types.

Our reduced core also includes a truncation operator $\lfloor \rfloor$ for
function applications, method applications, and the vararg expression.
We use $\lfloor a \rfloor_{0}$ to denote function and method applications
that produce no value, because they appear as statements.
We use $\lfloor me \rfloor_{1}$ to denote function applications,
method applications, and vararg expressions that produce only one value,
even if they return mutiple values.

@@

As an example, lets assume that $x$ and $y$ are variables in the
environment with types $\Integer$ and $\String$, the assignment
\[
x, y = 1, ``foo"
\]
type checks through the following rules:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env_{i}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{e}\;|}
      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env_{i}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\end{array}
\]

@@

Our previous example type checks through these rules because
\[
1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}
\]

As another example, lets assume that $x$, $y$, and $z$ are variables in
the environment with types $\Integer$, $\String$, and $\String \cup \Nil$,
the assignment
\[
x, y, z = 1, ``foo"
\]
type checks because
\[
1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
\]

Note how $\Nil{*}$ replaces any missing values.
This example type checks because $\Nil{*}$ produces as many $\Nil$
values as we need, and $\Nil$ is consistent with $\String \cup \Nil$,
which is the type of $z$.

Conversely, the assignment
\[
x = 1, ``foo"
\]
type checks because
\[
1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \Value{*}
\]

Note how $\Value{*}$ discards extra values.
This example type checks because $\Value{*}$ discards as many extra
values as we need, and $``foo"$ is consistent with $\Value$.

@@

As an example, assuming that $f$ is a local function in the environment, and that $f$ has type
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*} \rightarrow \Integer{*}$,
the function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]
and the function call
\[
f(``foo",1,2,3)
\]
also type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]

@@

We just mentioned that when our type system type checks an expression list,
it always includes $\Nil{*}$ in the end of the type of this expression list
if its type does not end in a variadic type.
This behavior preserves the semantics of Lua on replacing missing values,
and it is necessary when we omit optional parameters in a function call,
like the previous example showed.

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
We made this restriction because our type system does not keep
track of aliases to table fields.
This means that allowing \emph{unique} and \emph{open} table
types to appear in the type of a value would allow the
creation of unsafe aliases.
As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$.

After presenting some typing rules of the table constructor,
we start the discussion of the rules that define the most
unusual feature of our type system: the refinement of table types.
The first kind of refinement allows programmers to add new
fields to \emph{unique} or \emph{open} table types through
field assignment.
For instance, in Section \ref{sec:tables} we presented the
following example:
\begin{verbatim}
    local person = {}
    person.firstname = "Lou"
    person.lastname = "Reed"
\end{verbatim}

We can translate this example to our reduced core as follows:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; person = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$person[``firstname"] \; {<}\String{>} = ``Lou";$}\\
& \multicolumn{1}{l}{$person[``lastname"] \; {<}\String{>} = ``Reed"$}
\end{tabular}
\end{center}

@@

We use the refinement of table types to handle the declaration of
new global variables.
In Lua, the assignment \texttt{v = v + 1} translates to
\texttt{\string_ENV["v"] = \string_ENV["v"] + 1} when \texttt{v}
is not a local variable, where \texttt{\string_ENV} is a table
that stores the global environment.
For this reason, Typed Lua treats accesses to global variables as field accesses
to an \emph{open} table in the top-level scope.
In the following examples we assume that $\string_ENV$ is in the
environment and has type $\{\}_{open}$.

As an example,
\[
\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``y"] \; {<}\Integer{>} = 1
\]
uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
Therefore, after these field assignments $\string_ENV$ has type
$\{``x":\String, ``y":\Integer\}_{open}$.

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
As we mentioned in Sections \ref{sec:oop} and \ref{sec:subtyping},
we need \emph{fixed} table types to type classes in object-oriented programming.
In the implementation we fix the aliasing of \emph{unique} and \emph{open}
table types that appear in a return statement, and in other cases we
close the aliasing of \emph{unique} and \emph{open} table types.
However, in the formalization we chose to define this behavior not
deterministically, as it makes easier the presentation of this behavior.
Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env, \penv \vdash id:close(T_{2}), \env[id \mapsto open(T_{1})]}
\end{array}
\;
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env, \penv \vdash id:fix(T_{2}), \env[id \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

Both rules use the auxiliary function \emph{read} because they may be
accessing an identifier that is bound to a filter or projection type.
As we mentioned in Section \ref{sec:unions}, our type system includes
a small set of \texttt{type} predicates that allow programmers to
discriminate union types, and our type system uses filter and projection
types in the definition of these predicates to handle the discrimination
of unions types.
While filter types discriminate unions of first-level types, projection
types discriminate unions of second-level types and project unions of
second-level types into unions of first-level types.
We can define \emph{read} as follows:
\begin{align*}
read(\penv, \phi(T_{1},T_{2})) & = T_{2}\\
read(\penv, \pi_{i}^{x}) & = proj(\penv(x), i)\\
read(\penv, T) & = T
\end{align*}

@@

As an example,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{``foo":\Integer\}_{unique} = \{ [``foo"] = 5 \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{closed} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$b[``foo"] = \mathbf{nil}$}
\end{tabular}
\end{center}
does not type check,
as aliasing $a$ produces the type $\{``foo":\Integer\}_{closed}$
that is not a subtype of $\{``foo":\Integer \cup \Nil\}_{closed}$,
the type of $b$.
Our type system has this behavior to warn programmers about
potential unsafe behaviors after this kind of alias.
In this example, it is unsafe to assign $\mathbf{nil}$ to the field
$`foo"$ of $b$, as it changes the value that is stored in the field
$``foo"$ of $a$.

As another example,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
does not type check, as aliasing $a$ produces the type $\{\}_{closed}$
that is not a subtype of $\{\}_{open}$, the type of $b$.
In this example, it is unsafe to add the field $``x"$ to $b$,
as it changes the value that is stored in the field $``x"$ of $a$.

@@

As an example, rule \textsc{T-FUNCTION1} prevents the following unsafe example to type check:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Void \rightarrow \Integer \times \Void =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Void$}\\
& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])$}
\end{tabular}
\end{center}

@@

First, our type system uses rule \textsc{T-EXPLIST6} for type checking
$idiv(1, 2)$:
\[
\begin{array}{c}
\mylabel{T-EXPLIST6}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void \sqcup T_{1}' \times ... \times T_{n}' \times \Void, \env_{2}\\
       S = T_{1} \times ... \times T_{n} \times \Nil{*} \sqcup T_{1}' \times ... \times T_{n}' \times \Nil{*}
       \end{array}}
      {\env_{1}, \penv \vdash me:\pi_{1}^{x} \times ... \times \pi_{n}^{x} \times \Nil{*}, \env_{2}, (x,S)}
\end{array}
\]

Rule \textsc{T-EXPLIST6} uses the auxiliary relation
$\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
This relation means that given a type environment $\env_{1}$ and
a projection environment $\penv$, we can check that an expression
list $el$ has type $S_{1}$ and produces a new type environment $\env_{2}$
and produces a pair $(x,S_{2})$.
This pair means that the last expression of an expression list $el$
produces an union of second-level types $S_{2}$ that sould be bound
to a variable $x$ in the projection environment $\penv$,
as the resulting type of this expression is a tuple of projection
types $\pi_{i}^{x}$.
In our example, we can check that $idiv(1, 2)$ has type
$\pi_{1}^{x} \times \pi_{2}^{x} \times \Nil{*}$ and produces the pair
$(x,\Integer \times \Integer \times \Nil{*} \sqcup \Nil \times \String \times \Nil{*})$.

Our type system uses the pair $(x,S_{2})$ to allow rule
\textsc{T-LOCAL2} to bound a union of second-level types $S_{2}$
to a variable $x$ in the projection environment $\penv$:
\[
\begin{array}{c}
\mylabel{T-LOCAL2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash el:S_{1}, \env_{2}, (x,S_{2})\\
       \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)], \penv[x \mapsto S_{2}] \vdash s, \env_{3} \;\;\;
       n = |\;\overline{id}\;|  
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{3} - \{\overline{id}\}}
\end{array}
\]

This rule uses the auxiliary function \emph{infer} to get the
first-level types of each variable that should be introduced in
the type environment for type checking $s$.
After type checking the statement $s$, rule \textsc{T-LOCAL2} produces a
new type environment $\env_{3}$ without the variables that it introduced
before type checking $s$.
In our example, we can check that declaring $q$ and $r$
bounds the projection type $\pi_{1}^{x}$ to $q$ and
bounds the projection type $\pi_{2}^{x}$ to $r$, where the projection variable $x$ bounds to 
\[
\Integer \times \Integer \times \Nil{*} \sqcup \Nil \times \String \times \Nil{*}
\]
in the projection environment $\penv$. 
This means that $q$ uses $\pi_{1}^{x}$ to project its type into the
union type $\Integer \cup \Nil$, while $r$ uses $\pi_{2}^{x}$ to
project its type into the union $\Integer \cup \String$.

Introducing a projection variable $x$ in the projection environment allows our
type system to discriminate projection types $\pi_{i}^{x}$.
The rule \textsc{T-IF3} shows the case where our type system
discriminates a projection type based on the tag \texttt{nil}:
\[
\begin{array}{c}
\mylabel{T-IF3}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
       \env_{4} = openset(closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\]

Rule \textsc{T-IF3} uses the auxiliary functions \emph{fpt} and \emph{gpt}
to filter a projection $x$, affecting all variables that bind to the same projection.
For instance, our previous example type checks through rule \textsc{T-IF3},
because it uses the information provided by the projection type $\pi_{1}^{x}$,
which is the type of $q$, to make the rule \textsc{T-IF3} use the function call
\[
fpt(\Integer \times \Integer \times \Nil{*} \sqcup \Nil \times \String \times \Nil{*}, \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Integer \times \Integer \times \Nil{*}$ inside the $\mathbf{if}$ branch,
and the function call
\[
gpt(\Integer \times \Integer \times \Nil{*} \sqcup \Nil \times \String \times \Nil{*}, \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Nil \times \String \times \Nil{*}$ inside the $\mathbf{else}$ branch.
Thus, reading $q$ and $r$ projects $\pi_{1}^{x}$ to $\Integer$ and
$\pi_{2}^{x}$ to $\Integer$ inside the $\mathbf{if}$ branch,
but it projects $\pi_{1}^{x}$ to $\Nil$ and $\pi_{2}^{x}$ to $\String$
inside the $\mathbf{else}$ branch.
Outside the condition, $q$ and $r$ use the original projection, that is,
they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.
